---
title: "PML Course Project Report"
author: "Alexander Kolchin"
date: "Wednesday, September 17, 2014"
output: html_document
---

## Synopsis
My goal in this project is to predict the manner in which people did the exercise by using the provided data from accelerometers on the belt, forearm, arm, and dumbbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

More information about provided data and their source is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

After loading, exploring and cleaning provided testing data set I create prediction model by using random forest method.

I split provided training data set to two parts in order to be able to estimate out of sample error by cross validation.

Finally I predict values for provided test data set and use predicted values to submit for assignment.

```{r echo=FALSE, results='hide', message=FALSE}
library(data.table)
library(caret)
library(foreach)
library(randomForest)

set.seed(100)

```

## Load data
```{r}
ext.url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/";
load.data <- function(file.name) {
  if (!file.exists(file.name)) {
    download.file(paste(ext.url, file.name, sep = ""), file.name)
    } 
  
  read.csv(file.name, na.strings=c("#DIV/0!", "", "NA"))
  }

pml.training <- load.data("pml-training.csv")

```


## Split data to training and testing sets for future cross validation

Next I have to split provided data in my turn to two sets: training and testing. I will use training set to create model and I will use testing set in order to evaluate out of sample error.
``` {r}
inTrain <- createDataPartition(pml.training$classe, p=.60, list=FALSE);
training <- pml.training[inTrain,]
testing <- pml.training[-inTrain,]

```

## Explore training data
``` {r}
dim(training)

```

## Preprocess / Clean data

### remove irrelevant columns 
Some columns are obviously useless in order to make predictions, I have to remove them.
``` {r}
feature.set <- colnames(training[, -which(colnames(training) %in% c('X', 'user_name', 'raw_timestamp_part_1', 'raw_timestamp_part_2', 'cvtd_timestamp', 'new_window', 'num_window'))])
training <- training[,feature.set]
dim(training)
```

### remove columns with NAs
Next I remove all columns which contain NA values.
``` {r}
feature.set <- colnames(training[,colSums(is.na(training)) == 0])
training <- training[,feature.set]
dim(training)
```

## Build prediction model
Finally I create model by applying randomForest on cleaned training set of data.
I did experiment with other models, but did not choose them by reason of accuracy or speed.
```{r}
training <- training[,feature.set]
model <- randomForest(classe ~ .,data = training)
```


## Evaluate expected out of sample error (use cross-validation)
In order to evaluate out of sample error I apply the model on the testing data set and evaluate predicted values.
```{r}
confusionMatrix(predict(model, newdata=testing), testing$classe)

```

## Apply model to the test data set
At the end I apply model to provided test data set.
```{r}
pml.testing <- load.data("pml-testing.csv")
predictions <- predict(model,newdata=pml.testing)
predictions
```

### Generate  files to submit for assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
  }

pml_write_files(predictions)
```

## Conclusion
As expected all 20 predicted values are correct answers for all test data provided in this assignment.

